Multi-Core Model Checking and Maximum Satisfiability Applied to Hardware-Software Partitioning

### 2015-12-03 Docs/IJES/journal.tex
minor changes

\LRH{A. Trindade et~al.}

@@

\begin{abstract}
We present new alternative approaches to solve the hardware and software partitioning problem. First we use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT) in conjunction with a multi-core support using Open Multi-Processing to create four variants to solve the partitioning problem. In a nutshell, the multi-core SMT-based BMC approach allows initializing many verification instances based on the number of available processing cores. Each instance checks for a different optimum value until the optimization problem is satisfied. We implement our algorithms on top of the Efficient SMT-Based Context-Bounded Model Checker (ESBMC). Other approach integrate the maximum satisfiability solver $\nu$Z tool into ESBMC. $\nu$Z provides a portfolio of approaches for solving linear optimization problems over SMT formulas. We compare all our approaches to a state-of-the-art optimization tool (MATLAB). Experimental results show that there is no single optimization tool to solve all hardware-software partitioning benchmarks.
\end{abstract}

@@

Nowadays, with the strong development of embedded systems, the design phase plays an important role. At early stages, the design is split into separated flows: hardware and software. The partitioning decision process, which deals with decisions upon which parts of the application have to be designed in hardware (HW) and which one in software (SW), must be supported by any well-structured methodology. If there is no methodology support, a number of issues, {\it e.g.}, design flow interruptions, redesigns, and undesired iterations may affect the overall development process, the quality, and the life-cycle of the final system.

Starting at the 1990s, intensive research was performed in HW-SW partitioning, and several approaches proposed, as shown in~\cite{Arato2003} and~\cite{Mann2007}. In~\cite{Trindade2015} and~\cite{Trindade2016} was shown that it is possible to use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT), implemented in a tool called Efficient SMT-Based Context-Bounded Model Checker (ESBMC) as defined in~\cite{Cordeiro2012}, in order to perform HW-SW partitioning in embedded systems. In~\cite{Trindade2016} the original ESBMC algorithm was corrected and the right algorithm caused a loss in terms of performance, so a multi-core sequential approach was introduced in order to bring the performance back and turn it similar, in some cases, to conventional optimization tools that are used to perform hardware-software partitioning.

The present work extends those studies since there is a substantial improvement in terms of the SMT-based verification methods, which has been extended and exhauted with a multi-core parallel- and binary-search approaches, as well as, the integration of the Maximum SMT (MaxSMT) solver $\nu$Z into ESBMC, which is a state-of-art optimization tool based on SMT as defined in ~\cite{Bjorner2015}.

Here, we exploit the availability of multi-core processors and the tool OpenMP as defined in~\cite{Tang2009}; in particular, SMT-based verification methods are applied to the HW-SW partition problem in three different ways using a multi-core ESBMC approach with OpenMP: ESBMC-SS using a sequential-search (SS), ESBMC-PS using a parallel-search (PS), and ESBMC-PB using a binary-search (BS). Experimental results are compared to ILP (integral linear programming), GA (generic algorithms) in a multi-core version, and also to $\nu$Z, which supports only a single-core approach, according~\cite{Bjorner2015}. The ILP and GA algorithms are implemented with the optimization toolbox of Matlab, according~\cite{TheMathWorks2013}, while $\nu$Z is a built-in tool to the SMT solver Z3. All multi-core ESBMC approaches, together with $\nu$Z, are implemented with the ESBMC tool.

@@

Our experiments are based on a set of publicly available benchmarks. All benchmarks, tools, and results of our evaluation are available on a supplementary web page\footnote{http://esbmc.org/}.

@@

Optimization is the act of obtaining the best result ({\it i.e.}, the optimal solution) under given circumstances as defined in~\cite{Rao2009}. There is no single method available for efficiently solving all optimization problems, according~\cite{Rao2009}. The most well-known technique is linear programming, which is an method applicable for the solution of problems in which the objective function and the constraints appear as linear functions of the decision variables. A particular case of linear programming is ILP, in which the variables can assume just integer values. Eq.~\ref{linear-programming-problem} shows a typical linear programming problem, where $A$ and $b$ are vectors or matrixes that describe the constraints

@@

\vspace{2 mm}
In some cases, the time to find a solution using ILP is impractical. Even with the use of powerful computers, a problem can take hours before an optimal solution is reached. If the optimization problem is complex, some heuristics can be used to solve the same problem faster, according~\cite{Rao2009}, {\it e.g.}, those used in the GA. The only drawback is that the found solution may not be the global minimum or maximum. Alternatively, tools such as ESBMC and $\nu$Z can be used to solve optimization problems so that the global minimum or maximum solution is found. The following sections describe the main features of ESBMC and $\nu$Z tools.

@@

Among the recent model checking techniques, there is one that combines model checking with satisﬁability solving. This technique, known as bounded model checking (BMC), does a very fast exploration of the state space, and for some types of problems, it offers large performance improvements over previous approaches, as shown in~\cite{Biere2009}. In particular, BMC based on Boolean Satisfiability (SAT) has been introduced as a complementary technique to binary decision diagrams for alleviating the state explosion problem, according~\cite{Clarke2001}.

@@

Fig.~\ref{ESBMC-Architecture} shows the current ESBMC architecture, which consists of the C/C$++$ parser, GOTO Program, GOTO Symex, and SMT solver, according~\cite{Ramalho2013}. In particular, ESBMC compiles C/C$++$ code into equivalent GOTO-programs ({\it i.e.}, control-flow graphs) using a gcc-compliant style. GOTO-programs can then be processed by the symbolic execution engine, called GOTO Symex, where two recursive functions compute the constraints ($C$) and properties ($P$); finally, it generates two sets of equations ({\it i.e.},\:$C \land \neg P$), which are checked for satisfiability by an SMT solver.

@@

To optimize the CPU resources utilization without modifying the underlying SMT solver, the Open Multi-Processing (OpenMP) library, according~\cite{Dagum1998}, is used in this present work as a front-end for ESBMC. Fig.~\ref{ESBMC-Multi-core} shows our first approach called ESBMC sequential-search ``ESBMC-SS''.

@@

In terms of SMT-based verification, most related studies are restricted to present the model, its modification to programming languages ({\it e.g.}, C/C++ and Java), and the application to multi-thread algorithms or to embedded systems to check for program correctness. In~\cite{Ramalho2013} it presents a bounded model checker for C++ programs, which is an evolution of dealing with C programs, and~\cite{Cordeiro2012} uses ESBMC for embedded ANSI-C software. In~\cite{Trindade2015}, and~\cite{Trindade2016} it was proven that it is possible to use ESBMC to solve HW-SW partitioning in a single- and multi-core way but the first had performance issues that were correct by the second, which used only a sequential search to perform multi-core model checking. There are related studies focused on decreasing the verification time of model checkers by applying Swarm Verification, as shown in~\cite{Holzmann2011}, and modifications of internal search engines to add support for parallelism, as shown in~\cite{Holzmann2012}, but there is still the need for initiatives related to parallel SMT solver, according~\cite{Wintersteiger2009}.

Recently, the SMT solver Z3 has been extended to pose and solve optimization problems modulo theories, as shown in~\cite{Bjorner2015}. In particular, 
$\nu$Z tool offers substantial performance improvement in optimization problems, according~\cite{Bjorner2014} and ~\cite{Bjorner2015}. As an application example, \cite{Pavlinovic2015} propose an approach which considers all possible compiler error sources for statically typed functional programming languages and reports the most useful one subject to some usefulness criterion. The authors formulate this approach as an optimization problem related to SMT and use $\nu$Z to compute an optimal error source in a given ill-typed program. The approach described by \cite{Pavlinovic2015}, which uses MaxSMT solver $\nu$Z, shows a significant performance improvement if compared to previous SMT encodings and localization algorithms.

@@

\bibitem[\protect\citeauthoryear{Trindade \textit{et al.}}{2016}]{Trindade2016}
Trindade, A., Ismail, H. and Cordeiro, L. (2016) 'Applying Multi-Core Model Checking to Hardware-Software Partitioning in Embedded Systems (short paper)', {\it V Brazilian Symposium on Computing Systems Engineering (to appear)}

>>>>

%\LRH{A. Trindade et~al.}

@@

\begin{abstract}
We present new alternative approaches to solve the hardware and software (HW-SW) partitioning problem. First, we use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT) in conjunction with a multi-core support using Open Multi-Processing to create four variants to solve the partitioning problem. In a nutshell, the multi-core SMT-based BMC approach allows initializing many verification instances using different approaches based on the number of available processing cores. In particular, each instance checks for a different optimum value until the optimization problem is satisfied. We implement our algorithms on top of the Efficient SMT-Based Context-Bounded Model Checker (ESBMC). Additionally, we integrate the maximum satisfiability solver $\nu$Z tool into ESBMC, which provides a portfolio of approaches for solving linear optimization problems over SMT formulas. We compare all proposed approaches to a state-of-the-art optimization tool (MATLAB). Experimental results show that there is no single optimization tool to solve all HW-SW partitioning benchmarks, but ESBMC-$\nu$Z outperforms all parallel approaches proposed in this paper.
\end{abstract}

@@

Nowadays, with the strong development of embedded systems, the design phase plays an important role. At early stages, the design is split into separated flows: hardware (HW) and software (SW). The partitioning decision process, which deals with decisions upon which parts of the application have to be designed in hardware and which one in software, must be supported by any well-structured methodology. If there is no methodology support, a number of issues, {\it e.g.}, design flow interruptions, redesigns, and undesired iterations may affect the overall development process, the quality, and the life-cycle of the final system.

Starting at the $90$'s decade, intensive research was performed in HW-SW partitioning, and several approaches proposed, as shown in~\cite{Arato2003} and~\cite{Mann2007}. In~\cite{Trindade2015} and~\cite{Trindade2016} was shown that it is possible to use Bounded Model Checking (BMC) based on Satisfiability Modulo Theories (SMT), implemented in a tool called Efficient SMT-Based Context-Bounded Model Checker (ESBMC) as defined in~\cite{Cordeiro2012}, in order to perform HW-SW partitioning in embedded systems. In~\cite{Trindade2016}, the original ESBMC algorithm was improved in terms of performance, so a multi-core sequential approach was introduced, in order to speed up the performance and turn it similar, in some cases, to conventional optimization tools, which are used to perform HW-SW partitioning.

The present work thus extends those previous studies since there is a substantial improvement in terms of the SMT-based verification methods. In particular, we extend the work of~\cite{Trindade2015} and~\cite{Trindade2016} to exploit a multi-core parallel- and binary-search approaches, as well as, the integration of the Maximum SMT (MaxSMT) solver $\nu$Z into ESBMC, which is a state-of-art optimization tool based on SMT as defined in~\cite{Bjorner2015}.

We exploit the availability of multi-core processors and the tool OpenMP as defined in~\cite{Tang2009}; thus, we apply SMT-based verification methods to the HW-SW partition problem in three different ways using a multi-core ESBMC approach with OpenMP: ESBMC-SS using a sequential-search (SS), ESBMC-PS using a parallel-search (PS), and ESBMC-PB using a binary-search (BS). Experimental results are compared to ILP (integer linear programming), GA (generic algorithms) in a multi-core version, and also to $\nu$Z, which supports only a single-core approach, according to~\cite{Bjorner2015}. The ILP and GA algorithms are implemented with the optimization toolbox of Matlab, according to~\cite{TheMathWorks2013}, while $\nu$Z is a built-in tool to the SMT solver Z3. All multi-core ESBMC approaches, together with $\nu$Z, are implemented with the ESBMC tool.

@@

Our experiments are based on a set of publicly available benchmarks. All benchmarks, tools, and results of our evaluation are available on a supplementary web page \texttt{http://esbmc.org/}.

\subsection{Organization of this Work}

@@

Optimization is the act of obtaining the best result ({\it i.e.}, the optimal solution) under given circumstances as defined in~\cite{Rao2009}. There is no single method available for efficiently solving all optimization problems, according to~\cite{Rao2009}. The most well-known technique is linear programming, which is an method applicable for the solution of problems in which the objective function and the constraints appear as linear functions of the decision variables. A particular case of linear programming is ILP, in which the variables can assume just integer values. Eq.~\ref{linear-programming-problem} shows a typical linear programming problem, where $A$ and $b$ are vectors or matrixes that describe the constraints

@@

\vspace{2 mm}
In some cases, the time to find a solution using ILP is impractical. Even with the use of powerful computers, a problem can take hours before an optimal solution is reached. If the optimization problem is complex, some heuristics can be used to solve the same problem faster, according to~\cite{Rao2009}, {\it e.g.}, those used in the GA. The only drawback is that the found solution may not be the global minimum or maximum. Alternatively, tools such as ESBMC and $\nu$Z can be used to solve optimization problems so that the global minimum or maximum solution is found. The following sections describe the main features of ESBMC and $\nu$Z tools.

@@

Among the recent model checking techniques, there is one that combines model checking with satisﬁability solving. This technique, known as bounded model checking (BMC), does a very fast exploration of the state space, and for some types of problems, it offers large performance improvements over previous approaches, as shown in~\cite{Biere2009}. In particular, BMC based on Boolean Satisfiability (SAT) has been introduced as a complementary technique to binary decision diagrams for alleviating the state explosion problem, according to~\cite{Clarke2001}.

@@

Fig.~\ref{ESBMC-Architecture} shows the current ESBMC architecture, which consists of the C/C$++$ parser, GOTO Program, GOTO Symex, and SMT solver, according to~\cite{Ramalho2013}. In particular, ESBMC compiles C/C$++$ code into equivalent GOTO-programs ({\it i.e.}, control-flow graphs) using a gcc-compliant style. GOTO-programs can then be processed by the symbolic execution engine, called GOTO Symex, where two recursive functions compute the constraints ($C$) and properties ($P$); finally, it generates two sets of equations ({\it i.e.},\:$C \land \neg P$), which are checked for satisfiability by an SMT solver.

@@

To optimize the CPU resources utilization without modifying the underlying SMT solver, the Open Multi-Processing (OpenMP) library, according to~\cite{Dagum1998}, is used in this present work as a front-end for ESBMC. Fig.~\ref{ESBMC-Multi-core} shows our first approach called ESBMC sequential-search ``ESBMC-SS''.

@@

In terms of SMT-based verification, most related studies are restricted to present the model, its modification to programming languages ({\it e.g.}, C/C++ and Java), and the application to multi-thread algorithms or to embedded systems to check for program correctness. In~\cite{Ramalho2013} it presents a bounded model checker for C++ programs, which is an evolution of dealing with C programs, and~\cite{Cordeiro2012} uses ESBMC for embedded ANSI-C software. In~\cite{Trindade2015}, and~\cite{Trindade2016} it was proven that it is possible to use ESBMC to solve HW-SW partitioning in a single- and multi-core way, but the former has performance issues that were improved by the latter, which used only a sequential search to perform multi-core model checking. There are related studies focused on decreasing the verification time of model checkers by applying Swarm Verification, as shown in~\cite{Holzmann2011}, and modifications of internal search engines to add support for parallelism, as shown in~\cite{Holzmann2012}, but there is still the need for initiatives related to parallel SMT solver, according to~\cite{Wintersteiger2009}.

Recently, the SMT solver Z3 has been extended to pose and solve optimization problems modulo theories, as shown in~\cite{Bjorner2015}. In particular, $\nu$Z tool offers substantial performance improvement in optimization problems, according to~\cite{Bjorner2014} and~\cite{Bjorner2015}. As an application example, \cite{Pavlinovic2015} propose an approach which considers all possible compiler error sources for statically typed functional programming languages and reports the most useful one subject to some usefulness criterion. The authors formulate this approach as an optimization problem related to SMT and use $\nu$Z to compute an optimal error source in a given ill-typed program. The approach described by \cite{Pavlinovic2015}, which uses MaxSMT solver $\nu$Z, shows a significant performance improvement if compared to previous SMT encodings and localization algorithms.

@@

\bibitem[\protect\citeauthoryear{Trindade \textit{et al.}}{2015}]{Trindade2016}
Trindade, A., Ismail, H. and Cordeiro, L. (2015) 'Applying Multi-Core Model Checking to Hardware-Software Partitioning in Embedded Systems (short paper)', {\it V Brazilian Symposium on Computing Systems Engineering (to appear)}
